The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-29 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Small textualities

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

@@

% TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to integrate sufficiently long to reach a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, which is one magnitude away from expected EoR signals. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. Those analyses conclude that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, the leaked RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

In this paper, we will describe the mitigation efforts implemented for the MWA, and show examples of RFI found and results of the RFI detection. Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved. This strategy is applied to 63~h of MWA data, and these data are described in Sect.~\ref{ch:data-description}. Sect.~\ref{ch:detection-results} presents the RFI that was found in these data, as well as the effectivity of the RFI detection. In Sect.~\ref{ch:lofar-comparison}, these results will be compared to a previously performed LOFAR RFI survey. Finally, conclusions are drawn and discussed in Sect.~\ref{ch:conclusions-and-discussion}.

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast. It also has a library interface, which allows it to be integrated in a pipeline.

@@

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations. For this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observations except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy. TODO what are these transmitters?

@@

Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.

@@

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?. \textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity (e.g. TODO Miriad, eMERLIN), but other projects still use single-sample thresholding, e.g. PAPER. Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.

>>>>

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyze the impact of RFI on a specific instrument to optimize the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

@@

% TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and conclude in \citet{offringa-rfi-distributions} that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

In this paper, we will describe the mitigation strategy implemented for the MWA, and show examples of RFI found and results of the RFI detection. Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved. This strategy is applied to 63~h of MWA data. In Sect.~\ref{ch:data-description}, these data are described. Sect.~\ref{ch:detection-results} presents examples of RFI that were found in these data, as well as the effectivity of the RFI detection. In Sect.~\ref{ch:lofar-comparison}, the results are compared to a previously performed LOFAR RFI survey. Finally, in Sect.~\ref{ch:conclusions-and-discussion} conclusions are drawn and discussed.

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy. TODO what are these transmitters?

@@

Because significant DVB interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.

@@

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?. Second, \textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Finally, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity (e.g. TODO Miriad, eMERLIN), but other projects still use single-sample thresholding, e.g. PAPER. Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.
